---
title: "Chapter 5: Splines"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width = 5, fig.height = 3.2, dpi = 200)
```

## 5.1 Smoothing Splines

### Natural Cubic Splines

If we order our points in increasing value a natural cubic spline is a function 
made up of sections of cubic polynomials linking each successive pair of points.
Of all the functions that are continous and interpolate the points the
"smoothest" minimizes $\int_{x_{1}}^{x_{n}}f''(x)^{2}dx$.

### Cubic Smoothing Splines

Because most data is noisy and you don't have a full grasp of the
data-generating mechanism you usually want to smooth the data rather than
interpolate the points. So instead of fixing $g(x_i) = y_i$ we set them as
parameters and attempt to minimize 

\[
\sum_{i=1}^{n}(y_{i} - g(x_i))^2 + \lambda\int{g''(x)dx}
\]

with $\lambda$ as the tuning parameter. We call $g(x)$ a *smoothing spline*. 

While these are ideal smoothers (smoothing function that minimizes error) they have
as many free parameters as there are data so can be problematic to estimate.
Also because we are smoothing the function anyway we almost certainly won't need
all n parameters.

## 5.2 Penalized Regression Splines

A Regression spline constructs a spline basis and then uses that basis to 
model the original data set. 

## 5.3 Some One-Dimensional Smoothers

### Cubic Splines

Already talked about these. If I get adventurous I'll try to fit one manually

From `mgcv` these can be called with `s(x, bs = "cr")`

### Cyclic splines

These match the outermost knots so that the smooth function is equal there. 
Think of modeling smooth effects throughout the year, you wouldn't want there to
be a discontinuity at the end of the year.

### B-splines

A B-Spline (Basis Spline) are only non-zero between m + 3 adjacent knots (m + 1
is the order of the basis, so for cubic splines m = 2 and a B-spline would be
non-zero for the nearest 5 knots). This makes them stable to compute. We can
define them recursively by saying any m order spline is a weighted sum of lower
order splines where the weights are defined by how close the x-values are to
the series of knots. Wikipedia says:

> (the first) ramps from zero to one as x goes from t_{i} to t_{i+k} and (the second)
ramps from one to zero as x goes from t_{i+1} to t_{i+k+1}.

The final stage of recursion is just a binary indicating
which knot span x is in. 

So if we want a cubic B-spline with 6 knots we actually need to define 10 knots; 6
for the locations, 2 for the ends, and 2 for the cubic order. 

```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
x <- 1:100 + rnorm(100)
k <- seq(0, 100, length.out = 10)

print(x)
print(k)

bspline <- function(x, k, i, m = 2){
  ## Evaluate ith B-spline basis function of order m at the
  ## values in x, given knot locations in k
  if(m == -1){
    res <- as.numeric(x < k[i + 1] & x >= k[i])
  } else {
    z0 <- (x - k[i]) / (k[i + m + 2] - k[i + 1])
    z1 <- (k[i + m + 2] - x) / (k[i + m + 2] - k[i + 1])
    res <- z0 * bspline(x, k, i, m - 1) + z1 * bspline(x, k, i + 1, m - 1)
  }
  return(res)
}


spline_basis <- vapply(1:6, function(i, x_ = x, k_ = k) bspline(x = x_, k = k_, i = i), numeric(100))
spline_basis_df <- data.frame(spline_basis, stringsAsFactors = F)
names(spline_basis_df) <- as.character(seq_len(ncol(spline_basis)))
spline_basis_df$X <- x
  
spline_basis_df %>%
  gather(Function, Value, -X) %>%
  ggplot(aes(x = X, y = Value, color = Function)) + 
  geom_line(size = 2) +
  geom_vline(aes(xintercept = loc), data = data_frame(loc = k), linetype = "dashed") +
  ggtitle("B-Spline Representation", "dashed lines are knot locations") +
  theme_bw() +
  xlab("Original X Value") +
  ylab("Basis Transformation Value")
```

So as you can see the outside 2 knots on both sides don't actual have full
coverage, which is why we need to put knot locations outside the range of our
data. We would then use this new basis as the variables in our regression model. We
would then use these transformed values as inputs for our regression model. 

```{r}
knitr::kable(head(spline_basis_df), align = "c")
```

### P-Splines

P-splines are low rank smoothers using a B-spline basis, but with a *difference
penalty* applied to the parameters to control wiggliness. This means we are 
penalizing the squared differences between adjacent $\beta_i$ values. We can 
represent this penalty as 

\[
\sum_{i=1}^{k-1}(\beta_{i+1} - \beta_i)^2 = \mathbf{\beta^TP^TP\beta}
\]

where P is just a diagonal difference matrix. 

```{r}
k <- 6
P <- diff(diag(k), differences = 1)
S <- t(P) %*% P
S
```

P-Splines do require evenly spaced knots, but other than that they are very
flexible.

From `mgcv` these can be called with `s(x, bs = "ps", m = c(2, 3))` where m
is the order for the basis and penalties, respectively. 

### Adaptive Smoothers

Sometimes we want the amount of smoothing to vary along with the x value. 

From `mgcv` these can be called with `s(x, bs = "ad", m = c(2, 3))`



