---
title: "Survival Analysis Notes"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Cox Proportional Hazards Model

We observe event times along with covariates which may help to predict the event
times. We define a *hazard function*, $h(t)$, which gives the probability per
unit time of the event occuring at $t$, given that it has not already occured.
The Cox model attempts to measure the effects of the covariates on the hazard as
opposed to measuring $h(t)$ directly.

The model for the $i^{th}$ subject is

\[
h_{i}(t) = h_{o}(t)exp(\mathbf{X}_{i}(t)\beta)
\]

$h_{o}(t)$ is the baseline hazard so each subject's baseline hazard gets
modified by their coefficient values.

One thing to note is that the coefficient values are only estimated when an
event occurs. Not 100% on why this is. 

The Probability that the $i^{th}$ subject expired at $t_{i}$ is (if $R(i)$
is the set of all subjects at risk at time $t$)

\[
exp(\mathbf{X}_{i}\beta) / \sum_{j \in{R(i)}} exp(\mathbf{X}_{j}\beta)
\]

So the liklihood becomes

\[
\prod_{i} exp(\mathbf{X}_{i}\beta) / \sum_{j \in{R(i)}} exp(\mathbf{X}_{j}\beta)
\]

If subjectes leave the study before their event occurs (or if the study ends
before they do) then they still count in $R(i)$ (The denominator) but don't
exist in the numerator. This is called Right Censoring. 

Because we are not estimating $h_{o}$ (it cancels out of the above equations)
this is called the *partial liklihood*. 

### Cumulative Hazard and Survival Functions

The probability of surviving until time $t$ is called the *survival function*,
$S(t)$. If $H(t) = \int_{0}^{t}h(x)dx$ then $S(t) = exp(-H(t))$. The baseline
hazard can then be given as

\[
\hat{H}_{o}(t) = \sum_{t_{j} \leq t}exp(\hat{\alpha}_{j})
\]

where $\alpha_{i}$ is the intercept term for the $i^{th}$ time. 

Another way to think about this is the following equation:

\[
exp(\alpha_{j}) = \int_{t_{j-1}^{+}}^{t_{j}}{h_{o}(x)dx}
\]

Which basically says that the intercept at time $j$ is is the baseline hazard
function between the previous event and the current one. For a subject specific
hazard, $\hat{H}(t)$, you just add the subjects covariates to the intercept. 

The useful residuals are just $d_{i} - \hat{H}_{i}$

### Example

Fit a model using the `bone` data set; `t` is the time, `d` is a binary indicator 
where 1 is death and 0 is censored, and `trt` is a factor with levels `allo` and
`auto`. 

```{r, echo = F}
library(gamair)
data(bone)
knitr::kable(head(bone))
```

We are going to use the approach in the book of creating a record for every
event time a subject is alive for. 

```{r}
## order by t
bone$id <- 1:nrow(bone)
bone_ordered <- bone[order(bone$t), ]
## event times
et <- unique(bone_ordered$t[bone_ordered$d == 1])
## Starts of risk sets
es <- match(et, bone_ordered$t)
n <- nrow(bone_ordered)
# Times for risk sets
t <- rep(et, 1 + n - es)
print(et)
print(1 + n - es)
str(t)
st <- cbind(0,
            bone_ordered[unlist(apply(matrix(es), 1, function(x, n) x:n, n = n)),])
## Signal Events
st[(st$t == t) & (st$d != 0), 1] <- 1
## Reset Event Time to risk set time
st$t <- t
names(st)[1] <- "z"

knitr::kable(head(st))
pb <- st
pb$tf <- factor(pb$t)
bone_gam <- glm(z ~ tf + trt - 1, poisson, pb)
```

`tf` is first so it is a contrast free term, so we estimate a seperate $a_{j}$
for each event.

```{r}
cum_hazard <- tapply(fitted(bone_gam), pb$id, sum)
## Martingale Residuals
m_resid <- bone$d - cum_hazard
summary(bone_gam)
drop1(bone_gam, test = "Chisq")
```

So the `auto` treatment is not significant in the model. A Positive value for
the `trtauto` variable means that the hazard is estimated to be higher for the
`auto` group. 

```{r}
library(dplyr)
library(ggplot2)
base_plot_df <- data_frame(te = sort(unique(bone$t[bone$d == 1])),
                           tf = factor(te),
                           trt = bone$trt[1]) %>%
  bind_rows(data_frame(te = sort(unique(bone$t[bone$d == 1])),
                           tf = factor(te),
                           trt = bone$trt[20]))

base_plot_df <- base_plot_df %>%
  mutate(raw_preds = as.numeric(predict(bone_gam, base_plot_df)),
         cum_hazard = cumsum(exp(raw_preds)),
         survive = exp(-cum_hazard))

X <- model.matrix(~tf + trt - 1, base_plot_df)
J <- apply(exp(base_plot_df$raw_preds) * X, 2, cumsum)
se_raw <- diag(J %*% vcov(bone_gam) %*% t(J))^.5

base_plot_df <- base_plot_df %>%
  mutate(se = exp(-cum_hazard + se_raw),
         se_minus = exp(-cum_hazard - se_raw))

ggplot(aes(x = te, y = survive, color = trt), data = base_plot_df) + 
  geom_step(size = 2) +
  geom_step(aes(y = se), linetype = "dashed") +
  geom_step(aes(y = se_minus), linetype = "dashed") +
  xlab("Time\n(days)") +
  ylab("Survival Function\nS(t)") +
  ggtitle("Survival Function for Allo Treatment", "Dashed Line is +/- 1 SE") +
  scale_y_continuous(breaks = seq(0, 1, by = .2), limits = c(0, 1)) +
  scale_color_discrete(name = "Treatment") +
  theme(legend.position = "top") +
  theme_bw()
```

## Survival Analysis with MGCV

We have data for drug trials. `pbc` contains baseline measures for each patient
and `pbcseq` contains some time-varying values. 

```{r, echo = F}
library(survival)
library(mgcv)
data(pbc)
knitr::kable(head(pbc))
knitr::kable(head(pbcseq))
```

The disease takes 4 `stage` values, with 1 being subtle damage and 4 is 
cirrhosis. The weights vector provides the censoring information (0 for
censoring, 1 for event)

```{r, fig.width=8,fig.height=6}
pbc$status1 <- as.numeric(pbc$status == 2)
pbc$stage <- factor(pbc$stage)

pbc_gam <- gam(time ~ trt + sex + s(sqrt(protime)) + s(platelet) + s(age) + s(bili) + s(albumin),
          weights = status1,
          family = cox.ph,
          data = pbc)

anova(pbc_gam)
```

So the treatment is not significant

```{r}
plot(pbc_gam, pages = 1); plot(pbc_gam$linear.predictors, residuals(pbc_gam))
```

The wedge of points on the residual plot is from the censored observations and
is expected in survival analysis residual plots. 

```{r}
np <- 300
newd <- data.frame(matrix(0, np, 0))
for(n in names(pbc)) newd[[n]] <- rep(pbc[[n]][25], np)
newd$time <- seq(0, 4500, length = np)
fv <- predict(pbc_gam, newdata = newd, type = "response", se = T)
newd$fit <- fv$fit
newd$se_fit <- fv$se.fit
newd$se <- with(newd, se_fit / fit)
newd$se_pred <- with(newd, exp(log(fit) + se))
newd$se_pred_minus <- with(newd, exp(log(fit) - se))

ggplot(aes(x = time, y = fit), data = newd) + 
  geom_step() +
  geom_step(aes(y = se_pred), linetype = "dashed") +
  geom_step(aes(y = se_pred_minus), linetype = "dashed") +
  xlab("Time") +
  ylab("Survival Function\nS(t)") +
  ggtitle("Predicted Survival Function for Patient 25") +
  theme_bw()
```


